---
title: How frequently do MLB teams with best record in regular season become World
  Champions?
author: "D. Hernandez (@gmbeisbol)"
date: '2016-11-16'
output:
  html_notebook: default
  html_document: default
  pdf_document: default
  word_document: default
self_contained: no
---
<img src = "images/HomePlate GM_brown.png" width = "50">

### Some reasons for this

This year the [Chicago Cubs](http://www.cubs.com) became the MLB *World Series Champions*, defeating the Cleveland Indians in a heartbreaking 7th game of the serie.

In 2016 the Cubs were also the team with the best record (number of wins) in all the Major Leagues.

Contrary to European Soccer (where teams play just a regular season and the one with best record becomes the league champion), in baseball, it's doesn't seem to be so frequent that the team with the best record is always the ones that achieves the Commissioner's Trophy.
This can be due to the fact that MLB, as any other sports leagues in America, has a playoff phase, on which teams that clinched the postseason have to play series against other team and win it in order to advance to next phases.

That's why I was curious after the end of the 2016 World Series to know how frequently a MLB team with the best record in the regular season, which could lead to think that it was the best team in the season, becomes the World Champion.

I used the "*Lahman*" R package as it has data of all MLB teams by season and it could be what I need to answer this question and maybe find some other insights about this.
So, let's play with the data.

First I loaded the packages used in the analysis.
```{r loading, message=FALSE, warning=FALSE}
# Loading the packages for the analysis
library(Lahman)
library(tidyverse)
library(plotly)
library(readxl)
library(knitr)
library(ggthemes)
library(svglite)
```

### Data acquisition and manipulation

#### Data acquisition

Using the table called `Teams` (Yearly Stats and Standings) from the [Lahman package](http://lahman.r-forge.r-project.org/), I created a [*tibble*](https://blog.rstudio.org/2016/03/24/tibble-1-0-0/) with a subset of it and called it `TeamsStd` (from Teams Standings).
I selected just a few columns from the original table. You can check the meaning of each of the selected columns [here](http://lahman.r-forge.r-project.org/doc/).
```{r}
TeamsStd <- tbl_df(Teams) %>%
  select(yearID, lgID, divID, Rank, W, L, DivWin, WCWin, LgWin, WSWin, name, teamIDBR)
```

So, printing a summary of the new `TeamsStd` tibble, we have:
```{r}
summary(TeamsStd)
```

#### Data manipulation

Fom the summary, it can be seen that `TeamsStd` has info from 1871 up to 2015. Nevertheless, the [World Series started on 1903](https://en.wikipedia.org/wiki/World_Series), so let's subset `TeamsStd` to have the data from 1903 to 2015.
```{r}
TeamsStd <- TeamsStd %>%
  filter(yearID >= 1903)
```

Let's add other variables to the tibble with `Winning-Loss Percentage` (called WLP) and `Total of Games` (called G), and let's order the rows per year based on their `WLP`, in descending order.
```{r}
TeamsStd <- mutate(TeamsStd,
          WLP = W/(W+L),
          G = W+L) %>%
  arrange(yearID, desc(WLP))
```

Now that the tibble is ordered by year and `WLP`, let's add a new variable with the overall rank (called `OverallRank`) for each team on each regular season, defining number 1 to the team with the best record, 2 for the second one and so on.
```{r}
TeamsStd <- mutate(TeamsStd,
         OverallRank = ave(WLP, yearID, FUN = seq_along))
```

Now `TeamsStd` has the info up to 2015. As the Lahman package is not yet updated with the 2016 data, in order to include the last World Series, I had to look for the 2016 data into [baseball reference](http://www.baseball-reference.com/leagues/MLB/2016-standings.shtml), built an Excel file (called `T2016.xlsx`) with the same variables than `TeamsStd` and then bind them together.
```{r, include=FALSE}
T2016 <- tbl_df(read_excel("C:/Users/1328/Documents/R projects/Best_Team_Champ/data/T2016.xlsx"))
```

```{r}
TeamsStd <- rbind(TeamsStd, T2016)
```

Additionally, to prepare better `TeamsStd` for the analysis, let's modify some of the `classes` of the variables and give better names to some of them:
```{r}
TeamsStd$WLP <- as.numeric(TeamsStd$WLP)
TeamsStd$yearID <- as.integer(TeamsStd$yearID)
TeamsStd$Rank <- as.factor(TeamsStd$Rank)
TeamsStd$W <- as.integer(TeamsStd$W)
TeamsStd$L <- as.integer(TeamsStd$L)
TeamsStd$G <- as.integer(TeamsStd$G)
TeamsStd$OverallRank <- as.factor(TeamsStd$OverallRank)
TeamsStd <- TeamsStd %>% rename(Season = yearID, League = lgID, Team = name)
TeamsStd <- mutate(TeamsStd,
                   Era = ifelse(Season <= 1968, "1903-1968",
                            ifelse(Season >= 1969 & Season < 1994, "1969-1993",
                                   ifelse(Season >= 1994 & Season < 2012, "1994-2011", "2012+"))))
TeamsStd$Era <- as.factor(TeamsStd$Era)

```

Now `TeamsStd` has data from 1903 to 2016, which is the period of analysis I want, and from where the exploratory analysis will begin.
`TeamsStd` has now the following view (subset of the 2016 season).
```{r}
knitr::kable(TeamsStd %>% slice(2391:2400), caption = "Subset of TeamsStd")
```

### Exploratory Analysis

Let's start by printing again a summary of `TeamsStd`:
```{r}
summary(TeamsStd)
```

#### Some facts from the summary view:
The maximum wins a team has achieved in a single season is 116.
Let's see which team was and if it became World Champions that season.
```{r}
knitr::kable(TeamsStd %>%
  filter(W == 116) %>%
  select(Season, League, OverallRank, Team, G, W, L, WLP, LgWin, WSWin))
```

So, there was not only one team, but two with the highest number of won games during a regular season, from 1903 to 2016.
None of both teams could became the World Champions those seasons, as shown in the previous table `WSWin = N`.

Curiously, the Chicago Cubs played the World Series `LgWin = Y` but they lost against the Chicago White Sox on 1906.

#### Which are the teams with best record in the regular season and that have won the World Series, since 1903?

To do this, I took a subset of the `TeamsStd` with the teams with the best `WLP` (best record) that became the World Champions (`WSWin = Y`) and save it as a new tibble, called `Best_Champs`.
```{r}
Best_Champs <- TeamsStd %>%
  filter(OverallRank == 1 & WSWin == "Y")
sort(table(Best_Champs$Team), decreasing = TRUE)
```

It seems that the New York Yankees are used to this kind of seasons, as 17 out of their [27 World Series](https://en.wikipedia.org/wiki/List_of_World_Series_champions) titles have occurred when they were the best team in the regular season. 
The following bar plot shows all these teams with their total of championships in that condition. This has occurred 48 times since 1903.

```{r dev="svg", echo=FALSE, fig.height=5, fig.width=8}
Best_Champs <- within(Best_Champs,
                       Team <- factor(Team,
                                      levels = names(sort(table(Team), decreasing = TRUE))))

brt_WS <- ggplot(Best_Champs, aes(x = Team)) +
  geom_bar(stat = "count", fill = "navajowhite4") +
  geom_text(stat="count", aes(label=..count..), vjust=-.5, size = 3, colour = "gray64") +
  labs(title = "World Champions Teams with best record",
       subtitle = "during regular season, since 1903",
       caption = "Data from Lahman R package 5.0-0",
       x = "Teams",
       y = "Championships") +
  theme_tufte() +
  theme(axis.text.x=element_text(angle = 60, hjust = 1, vjust = 1, size = 9),
        axis.text.y=element_blank(),
        axis.ticks = element_blank())

brt_WS

```

The bar plot clearly shows the dominance of the Yankees within this group. Indeed, no other team is near to achieve that level in the short term.
On the other hand the Chicago Cubs, with their 2016 championship, has made this (World Champion after having the best record) for the third time in their history (since 1903).

Note that the [Boston Americans](http://www.baseball-reference.com/teams/BOS/) became the Boston RedSox in 1907. Since 1958, the [Brooklyn Dodgers](http://www.baseball-reference.com/teams/LAD/) are the LA Dodgers, and the [NY Giants](http://www.baseball-reference.com/teams/SFG/) are the SF Giants.
The Oakland Athletics were the old [Philadelphia Athletics](http://www.baseball-reference.com/teams/OAK/) from 1901 to 1954.

So redefining the tibble to show the current Franchises, the bar plot can be reorganized to show the total of championships by  current Franchises:
```{r dev='svg', echo=FALSE, fig.height=5, fig.width=8}
Best_Champs$Team <- as.character(Best_Champs$Team)
Best_Champs$teamIDBR <- as.character(Best_Champs$teamIDBR)

Best_Champs <- mutate(Best_Champs,
                      Franchise = ifelse(Team == "Philadelphia Athletics" | Team == "Oakland Athletics", "Oakland Athletics",
                                         ifelse(Team == "Boston Americans" | Team == "Boston Red Sox", "Boston Red Sox",
                                               ifelse(Team == "New York Giants", "San Francisco Giants",
                                                      ifelse(Team == "Brooklyn Dodgers", "Los Angeles Dodgers",
                                                             Team)
                                                      )
                                               )
                                         ),
                      FranchID = ifelse(teamIDBR == "PHA" | teamIDBR == "OAK", "OAK",
                                         ifelse(teamIDBR == "NYG", "SFG",
                                                      ifelse(teamIDBR == "BRO", "LAD",
                                                             teamIDBR)
                                               )
                                        )
                      )

Best_Champs <- within(Best_Champs,
                       Franchise <- factor(Franchise,
                                      levels = names(sort(table(Franchise), decreasing = TRUE))))
brt_WS <- ggplot(Best_Champs, aes(x = Franchise)) +
  geom_bar(stat = "count", fill = "navajowhite4") +
  geom_text(stat="count", aes(label=..count..), vjust=-.5, size = 3) +
  labs(title = "World Champion Franchises with best record",
       subtitle = "during regular season, since 1903",
       caption = "Data from Lahman R package 5.0-0",
       x = "Teams",
       y = "Championships") +
  theme_tufte() +
  theme(axis.text.x=element_text(angle = 60, hjust = 1, vjust = 1, size = 9),
        axis.text.y=element_blank(),
        axis.ticks = element_blank())
brt_WS
```

Boston Red Sox, Oakland Athletics, Cincinnati Reds and Chicago Cubs sum up the same 17 titles of the NY Yankees on those conditions. 

#### When do those 48 championships occurred since 1903?
```{r, echo=FALSE, fig.height=2, fig.width=8}
Best_Champs <- within(Best_Champs,
                       FranchID <- factor(FranchID,
                                      levels = names(sort(table(FranchID)))))

brt_WS_years <- ggplot(Best_Champs, aes(x = Season, text = Team)) +
  geom_point(aes(y=""), shape = 23, size = 1.5, stroke = 0.2, fill = "blue", color = "orange", alpha = 0.8) +
  #geom_bar(stat = "count", fill = "blue", color = "white", show.legend = FALSE, width = 1) +
  labs(title = "Seasons when current franchises became Champions",
       subtitle = "with best record in regular season, since 1903",
       caption = "Data from Lahman R package 5.0-0",
       x = "Season",
       y = NULL) +
  theme_tufte() +
  theme(axis.ticks = element_blank(),
        panel.grid.major = element_line(colour = "darkgray", size = 0.2),
        panel.grid.major.y = element_blank()) +
  #      panel.grid.major.y = element_blank(),
  #     panel.grid.minor.y = element_blank(),
  #      panel.grid.minor = element_line(colour = "lightgray", linetype = "dotted", size = 0.1),
  #      panel.grid.major = element_line(colour = "darkgray", size = 0.2)) +
  scale_x_continuous(minor_breaks = seq(1900, 2015, 2), breaks = seq(1900, 2010, 10))
brt_WS_years
```

This plot suggest that this occurred more frequently between 1930 and 1950 than in the last decades. For example in the current century it has occurred just four times, including this year.

This could have some relation with the different format of postseasons along the history. Since 1903, [four different postseasons formats has been implemented](https://en.wikipedia.org/wiki/Major_League_Baseball_postseason#Format_history), where more rounds has been added in the postseason along the time, when the leagues were expaned with more teams.

##### Era 1903-1968:
For example, between 1903 and 1968, just one postseason round was played between the American League and National League Champions. That round was the itself the World Series.

On this format, if considering that both teams had evenly chances to win the *World Series (WS)*, theoretically 50% of the time the team with best record should became the World Champion.

##### Era 1969-1993:
From 1969 to 1993 a new round, previous to the World Series was added to the postseason. That round was called the *League Championship Series (LCS)*, and it was because both leagues expanded to twelve teams each and the leagues were subdivided into two divisions, the West and Central. The LCS was played, by league, between the teams with first place in the West Division and East Division.

On this format, if considering that the four teams had evenly chances to win the LCS and WS, theoretically 25% of the time the team with best record should became the World Champion (as the team with best record would have 50% of wining the LCS and 50% of winning the WS, thus as both events are independent, the probabilities are multipled).

##### Era 1994-2011:
On this Era the leagues were divided into three divisions, West, East and Central. Nevertheless, using only the divisional winners to advance to the postseason would make an odd number of teams in each league; so a fourth team was needed to rectify that situation. Then it was when the *Wild Card (WC)* team born in the MLB, allowing the second-best team on each league make the postseason, even if it was not a Division Champion.

This provoked that four teams on each league joined the postseason and a new round of was needed to define the two teams that would playe the LCS. This new round was called the *Division Series (DS)*.

On this format, if considering that the eight teams had evenly chances to win the DS, LCS and WS, theoretically 12.5% of the time the team with best record should became the World Champion (as in the previous era, adding a new round provoked the team with best record to play an addional series, so the team with best record would have 50% of changes to win the DS, 50% of winning the LCS, and 50% of winning the WS, thus as all three events are independent, the probabilities are also multipled).

##### Era 2012 to present:
This is the current postseason format and the difference with the 1994-2011 era is that an additional round was added only for the Wild Card teams, where the two teams qualify as WC, they play a one-game playoff and the winner advances to the DS.

On this format the team with best record does not have to play an additional round, so its probability reamins the same, 12.5% as in the previous era.

The following image summarizes these formats: ### ABC

So let's plot again the previous graph but differentiating the Eras:
```{r, echo=FALSE, fig.height=5, fig.width=8}

brt_WS_years <- ggplot(Best_Champs, aes(x = Season, text = Team)) +
  geom_point(aes(y = FranchID), shape = 23, size = 1.5, stroke = 0.2, fill = "blue", color = "orange", alpha = 0.8) +
  #geom_path()
  labs(title = "Seasons when current franchises became Champions",
       subtitle = "with best record in regular season, since 1903",
       caption = "Data from Lahman R package 5.0-0",
       x = "Seasons",
       y = NULL) +
  theme_tufte() +
  geom_vline(xintercept = c(1903, 1969, 1994, 2012), color = "red", alpha = 0.4, size = 1.5) +
  theme(axis.ticks = element_blank(),
        panel.grid.major = element_line(colour = "darkgray", size = 0.2),
        panel.grid.major.y = element_blank()) +
  scale_x_continuous(minor_breaks = seq(1900, 2015, 2), breaks = seq(1900, 2010, 10)) +
  annotate(geom = "text", x = 1935, y = "CHW", label = "Era 1903-1968", size = 3, color = "red") +
  annotate(geom = "text", x = 1981, y = "CHW", label = "1969-1993", size = 3, color = "red") +
  annotate(geom = "text", x = 2003, y = "CHW", label = "1994-2011", size = 3, color = "red") +
  annotate(geom = "text", x = 2016, y = "CHW", label = "2012+", size = 3, color = "red")

brt_WS_years
```

```{r, include=FALSE}
# Determine how many WS has been celebrated since 1903
WS <- TeamsStd %>%
          filter(WSWin == "Y")

# Determine how many WS has been won by a team with the best record in the regular season
Champs_BR <- TeamsStd %>%
  filter(WSWin == "Y" & OverallRank == 1)

length(Champs_BR$Season) / length(WS$Season) * 100

```
From 1903 to 2016, 112 World Series has been celebrated between the National and the American League Champions (there were no WS in 1904 and 1994). **That's equivalent to say that the best team in a regular season became World Champion in 43% of the time (48 out of 112).**.

Checking that situation based on the Era, we have:
```{r, fig.height=5, fig.width=8}
Best_Champs$frID_Era <- with(Best_Champs, paste(FranchID, Era, sep = "_"))


  
Best_Champs <- within(Best_Champs,
                       frID_Era <- factor(frID_Era,
                                      levels = names(sort(table(frID_Era), decreasing = TRUE))))

brt_WS <- ggplot(Best_Champs, aes(x = frID_Era)) +
  geom_bar(stat = "count", fill = "navajowhite4") +
  geom_text(stat="count", aes(label=..count..), vjust=-.5, size = 3) +
  labs(title = "World Champion Franchises with best record",
       subtitle = "during regular season, by Era",
       caption = "Data from Lahman R package 5.0-0",
       x = "Teams",
       y = "Championships") +
  theme_classic() +
  facet_grid(. ~ Era, scale = "free", space = "free") +
  theme(axis.text.x=element_text(angle = 90, hjust = 0, vjust = 0, size = 9),
        axis.text.y=element_blank(),
        axis.ticks = element_blank()) +
  scale_x_discrete("Teams", breaks = Best_Champs$frID_Era, label = Best_Champs$FranchID)
  #geom_text(label=Best_Champs$FranchID,angle=90,size=3)+
brt_WS
```

So, in the last two formats (Wild Card Eras), only five times it has occurred being made by only three teams.

```{r, echo=FALSE}
knitr::kable(Best_Champs %>%
  filter(Season >= 1994) %>%
  select(Season, League, OverallRank, Team, W, L, WLP), caption = "World Champions with best record in the Wild Card era")

```


```{r, fig.height=6, fig.width=8}

TeamsStd <- mutate(TeamsStd,
                      FranchID = ifelse(teamIDBR == "ANA" | teamIDBR == "CAL" | teamIDBR == "LAA", "LAA",
                                        ifelse(teamIDBR == "FLA" | teamIDBR == "MIA", "MIA",
                                               ifelse(teamIDBR == "MON" | teamIDBR == "WAS" | teamIDBR == "WSN", "WSN",
                                                      ifelse(teamIDBR == "TBD" | teamIDBR == "TBR", "TBR",
                                                             teamIDBR)
                                               )
                                        )
                      ),
                      clinch_ps = ifelse((DivWin == "Y" | WCWin == "Y") & WSWin == "N", "Clinched Playoff",
                                      ifelse(WSWin == "Y", "World Champion", "Did not go to postseason"))
)
  
                   
TeamsOverall <- TeamsStd %>%
  filter(Season >= 1995)

TeamsOverall$OverallRank <- as.integer(TeamsOverall$OverallRank)

Overall_Linegraph <- ggplot(TeamsOverall, aes(x = Season, y = OverallRank)) +
  geom_line(color = "navajowhite4", size = .8) +
  geom_point(aes(color = clinch_ps, shape = clinch_ps)) +
  guides(color = FALSE) +
  scale_colour_manual(breaks = c("Clinched Playoff", "World Champion", "Did not go to postseason"),  
                     values = c("darkblue", "darkblue", "navajowhite4")) +
  scale_shape_manual(c("Clinched Playoff", "World Champion", "Did not go to postseason"),  
                     values = c(21, 19, 1)) +
  scale_y_reverse(breaks = c(1,30)) +
  facet_wrap(~ FranchID, ncol = 5) +
  labs(title = "Overall rank of MLB teams in regular season",
       subtitle = "based on WLP in Wild Card Eras (since 1995)",
       caption = "Data from Lahman R package 5.0-0")+
  theme_tufte() +
  theme(axis.ticks = element_blank(),
        panel.grid.major.y = element_line(colour = "gray", linetype = "dotted", size = 0.1),
        panel.grid.minor.y = element_blank(),
          #panel.grid.minor = element_line(colour = "gray", linetype = "dotted", size = 0.3),
          #panel.grid.major = element_line(colour = "gray", linetype = "dotted", size = 0.1),
        strip.text.x = element_text(size = 10, family = "serif", face = "bold", colour = "black", angle = 0),
        axis.text.x=element_text(angle = 90, hjust = 0, vjust = 1, size = 7),
        axis.text.y=element_text(angle = 0, hjust = 1, vjust = 0.5, size = 6)) +
  scale_x_continuous(breaks = seq(1995, 2020, 5))

Overall_Linegraph
```
svglite("Test.svg")
Overall_Linegraph
dev.off()
